### 2011-09-08 paper.tex
Added Newick representation

$\startsymbol$

@@

\noindent
\begin{tabular}{ll}
Symbol & Meaning \\
\hline
$\grammar$ & Grammar \\
$\nodelabels$ & Set of node labels \\
$\terminals$ & Set of terminals \\
$\rules$ & Set of transformation rules \\
$\weight$ & Rule weight function \\
$\epsilon$ & The empty string \\
$\terminals^\ast$ & Set of strings over $\terminals$, including the empty string \\
$\outseq$ & Output sequence \\
$|\outseq|$ & Length of $\outseq$ \\
$\outsubseq{i}{j+1}$ & Substring of $\outseq$ from $i$ to $j$ inclusive ($i$ starts at 1) \\
$\outsubseq{i}{i}$ & The empty string \\
$\m(i,j,k,l,X)$ & Max parse tree weight for $\outsubseq{k}{l}$ rooted at $X[\outsubseq{i}{j}]$ \\
$\Delta(i,j,x)$ & Indicates if rule string $x$ matches output string $\outsubseq{i}{j}$ \\
$\Delta(i,x)$ & Indicates if rule string $x$ matches output string $\outseq$, starting at position $i$ \\
\end{tabular}

>>>>

\Tree[ .$\epsilon$ [ .$\startsymbol$ [ .$\epsilon$ ] ] ]

Each ``intermediate tree'' is, formally, an ordered tree whose nodes are labeled from
$(\nodelabels \cup \terminals^\ast)$.

The transformation rules can take the various forms shown in \tabref{RuleTypes}.

\begin{table}
\tiny
\begin{tabular}{m{.2in}m{1in}m{.2in}m{1in}r}
\centerline{Type} & \centerline{From} & & \centerline{To} & Newick representation
\\ \hline
(1) &
\Tree[ .$\alpha$ [ .$A$ [ .$\beta$ ] ] ]
 & $\to$ & 
\Tree[ .$\alpha$ [ .$B$ [ [ [ .$C$ $\epsilon$ ] ] [ [ .$\beta$ ] ] ] ] ]
& $((\beta)A)\alpha \to ((C,\beta)B)\alpha$
\\ \hline
(2) &
\Tree[ .$\alpha$ [ .$A$ [ .$\beta$ ] ] ]
 & $\to$ & 
\Tree[ .$\alpha$ [ .$B$ [ [ [ .$\beta$ ] ] [ [ .$C$ $\epsilon$ ] ] ] ] ]
& $((\beta)A)\alpha \to ((\beta,C)B)\alpha$
\\ \hline
(3) &
\Tree[ .$\alpha$ [ .$A$ [ .$\beta$ ] ] ]
 & $\to$ & 
\Tree[ .$\alpha$ [ [ [ .$C$ $\epsilon$ ] ] [ [ .$B$ $\beta$ ] ] ] ]
& $((\beta)A)\alpha \to (C,((\beta)B))\alpha$
\\ \hline
(4) &
\Tree[ .$\alpha$ [ .$A$ [ .$\beta$ ] ] ]
 & $\to$ & 
\Tree[ .$\alpha$ [ [ [ .$B$ $\beta$ ] ] [ [ .$C$ $\epsilon$ ] ] ] ]
& $((\beta)A)\alpha \to (((\beta)B),C)\alpha$
\\ \hline
(5) &
\Tree[ .$\alpha$ [ .$A$ [ .$\beta$ ] ] ]
 & $\to$ & 
\Tree[ .$\alpha$ [ .$B$ [ .$C$ $\beta$ ] ] ]
& $((\beta)A)\alpha \to (((\beta)C)B)\alpha$
\\ \hline
(6) &
\Tree[ .$\alpha$ [ .$A$ [ .$\beta$ ] ] ]
 & $\to$ & 
\Tree[ .$\alpha$ [ $u$ [ .$B$ [ $v$ [ $\beta$ ] $w$ ] ] $x$ ] ]
& $((\beta)A)\alpha \to (u,((v,\beta,w)B),x)\alpha$
\\ \hline
\end{tabular}
\normalsize
\caption{
\tablabel{RuleTypes}
Types of transformation rule (i.e. tree adjunction rule) used in this paper.
Here $\alpha,\beta$ represent any subtree;
$A \in \nodelabels$ is the source node label;
$B,C \in (\nodelabels \cup \{\epsilon\})$ are the destination node labels;
$\epsilon$ is the empty string;
and
$u,v,w,x \in \terminals^\ast$ are (possibly empty) terminal strings.
}
\end{table}

The derivation stops when no further transformations can be applied.
The trees generated by this process have the property that every leaf node is labeled with a terminal sequence,
while every internal node is labeled either with $\epsilon$, or with a member of $\nodelabels$
which never appears on the left-hand side of a rule in $\rules$,
and to which no transformations can therefore be applied.

The final column of the above table shows a shorthand representation of each rule in Newick format,
probably the most widely-understood bioinformatics format for representing tree structures.
The initial tree, in this representation, is $((\epsilon)S)\epsilon$.
In the table, we have omitted the placeholder $\epsilon$'s at leaf nodes,
so that (for example) the rule
\[
((\beta)A)\alpha\ \to\ ((C,\beta)B)\alpha
\]
should strictly be read as
\[
((\beta)A)\alpha\ \to\ (((\epsilon)C,\beta)B)\alpha
\]

\newcommand\rulesubset[1]{\rules_{#1}}

Let $\rulesubset{n} \subseteq \rules$ denote the subset of rules of type $n$ according to \tabref{RuleTypes}.

@@

\noindent
\begin{tabular}{ll}
Symbol & Meaning \\
\hline
$\grammar$ & Grammar \\
$\nodelabels$ & Set of node labels \\
$\terminals$ & Set of terminals \\
$\rules$ & Set of transformation rules \\
$\rulesubset{n}$ & Set of transformation rules of type $n$, according to \tabref{RuleTypes} \\
$\weight$ & Rule weight function \\
$\epsilon$ & The empty string \\
$\terminals^\ast$ & Set of strings over $\terminals$, including the empty string \\
$\outseq$ & Output sequence \\
$|\outseq|$ & Length of $\outseq$ \\
$\outsubseq{i}{j+1}$ & Substring of $\outseq$ from $i$ to $j$ inclusive ($i$ starts at 1) \\
$\outsubseq{i}{i}$ & The empty string \\
$\m(i,j,k,l,X)$ & Max parse tree weight for $\outsubseq{k}{l}$ rooted at $X[\outsubseq{i}{j}]$ \\
$\Delta(i,j,x)$ & Indicates if rule string $x$ matches output string $\outsubseq{i}{j}$ \\
$\Delta(i,x)$ & Indicates if rule string $x$ matches output string $\outseq$, starting at position $i$ \\
\end{tabular}
